
##Health and Economic Impact of Weather Events in the US between in the 21st century

#####Daniel López Urbano
######January 25, 2015


####Abstract

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

In this report we will explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

###Data Processing

First we need to download and load the data from NOAA.

```{r DataLoading}

#Function to download the data -------------------------------------
read_data <- function(fileName, source_url) {  
  if (!file.exists(fileName)) {
    download.file(source_url, destfile = fileName, method = "curl")
  }
  dataset <- read.csv(fileName, stringsAsFactors=FALSE)
  dataset
}
#-------------------------------------------------------------------

#Downloading and Loading the Data
stormData <- read_data("repdata-data-StormData.csv.bz2", 
                       "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")

data_dim <- dim(stormData) #Dimensions of the Data
evtype_afer <- length(unique(stormData$EVTYPE)) #985 unique event types

```

The data contains `r data_dim[1]` records and `r data_dim[2]` variables. 

For the purpose of this study we have special interest in the variable 'EVTYPE', which tells us the type of weather event for each record. We find `r evtype_afer` unique event types, which is rather a high number. We need to do some cleaning.


There's no a standard format to name the events and we can find cases like `Frost/Freeze`, `FROST/FREEZE` and `FROST\\FREZE` which obviously refer to the same type of event. --So we convert to lower case and remove punctuations.--

```{r Cleaning}

#---- Cleaning the Data ---------
events <- tolower(stormData$EVTYPE) #all characters to lowercase
events <- gsub("[[:blank:][:punct:]+]", " ", events) #cleaning punct. characters

```

We can find several event types containing the word 'tornado' or 'flood'.
Queremos hacer el analisis en terminos generales, sin considerar por separado diferentes tipos de tornado, asi que haremos limpieza de los datos.

```{r EVTYPEGrouping}

events <- gsub(".*tornado.*", "tornado", events)
events <- gsub(".*flood.*", "flood", events)
events[ (grepl(".*tstm.*",events) | grepl(".*thunderstorm.*",events))  & grepl(".*wind.*",events)] <- "thunderstorm wind"
events <- gsub(".*hurricane.*", "hurricane", events)
events <- gsub(".*drought.*", "drought", events)
events <- gsub(".*heat.*", "heat", events)
events <- gsub(".*hail.*", "hail", events)
events[grepl(".*ice.*",events) | grepl(".*snow.*",events)] <- "snow and ice"

#number of unique event types after cleaning
evtype_before <- length(unique(events))

#update the dataframe
stormData$EVTYPE <- events

```
Now we have `r evtype_before` unique event types, and is time to sumarize the data we need.

```{r}
#---- Sumarize Data ---------
library(plyr)

### Health Damage ###

#Calculating the damage to population health that causes each event type
casualities <- ddply(stormData, .(EVTYPE), summarise,
                    FATAL = sum(FATALITIES),
                    INJUR = sum(INJURIES))

#Top 10 of the most harmful events
health_top10 <- head(casualities[order(casualities$FATAL + casualities$INJUR, decreasing=T),] , 10)
health_top10_names <- health_top10$EVTYPE

#Preparing the ggplot
fatalities_p <- health_top10[,1:2]
names(fatalities_p) <- c("EVENT","CASUALITIES")
fatalities_p$TYPE <- "Fatalities"

injuries_p <- health_top10[,c(1,3)]
names(injuries_p) <- c("EVENT","CASUALITIES")
injuries_p$TYPE <- "Injuries"

health_plot <- rbind(fatalities_p , injuries_p)
health_plot$EVENT <- factor(health_plot$EVENT , levels = health_top10_names)
health_plot <- health_plot[order(health_plot$EVENT, decreasing=T),]

### Economic loss ###

#We need to convert the PROPDMGEXP and CROPDMGEXP field into a
#usuable factor and calculate the true damage to crops and property.

stormData$PROPDMGEXP<-toupper(stormData$PROPDMGEXP)
stormData$CROPDMGEXP<-toupper(stormData$CROPDMGEXP)

stormData[stormData$PROPDMGEXP %in% c("",NA, "-","?", "+"),]$PROPDMGEXP <- 0
stormData[stormData$PROPDMGEXP %in% c("H", "B"),]$PROPDMGEXP <- 2
stormData[stormData$PROPDMGEXP %in% "K",]$PROPDMGEXP <- 3
stormData[stormData$PROPDMGEXP %in% "M",]$PROPDMGEXP <- 6

stormData[stormData$CROPDMGEXP %in% c("",NA,"-","?", "+"),]$CROPDMGEXP <- 0
stormData[stormData$CROPDMGEXP %in% c("H", "B"),]$CROPDMGEXP <- 2
stormData[stormData$CROPDMGEXP %in% "K",]$CROPDMGEXP <- 3
stormData[stormData$CROPDMGEXP %in% "M",]$CROPDMGEXP <- 6

stormData$PROPDMG <- stormData$PROPDMG * 10 ** as.numeric(stormData$PROPDMGEXP)
stormData$CROPDMG <- stormData$CROPDMG * 10 ** as.numeric(stormData$CROPDMGEXP)

#Economic loss of each event type
econ_loss <- ddply(stormData, .(EVTYPE), summarize,
                   PROP = sum(PROPDMG),
                   CROP = sum(CROPDMG))
#top 10 events of economic loss
econ_top10 <- head(econ_loss[order(econ_loss$PROP + econ_loss$CROP, decreasing=T),] , 10)
econ_top10_names <- econ_top10$EVTYPE

```

With we've got so far, we can achieve our goal of evaluate the impact 

```{r TimeSeries}

#---- Time series Data --------

stormTimeLine <- stormData[stormData$EVTYPE %in% econ_top10_types[1:3], c('EVTYPE','PROPDMG','CROPDMG','BGN_DATE' )]
stormTimeLine$DATE <- as.Date(stormTimeLine$BGN_DATE, '%m/%d/%Y')
stormTimeLine$DMG <- stormTimeLine$PROPDMG + stormTimeLine$CROPDMG
stormTimeLine <- stormTimeLine[,c('EVTYPE','DMG','DATE')]

#Order by date
stormTimeLine <- stormTimeLine[order(stormTimeLine$DATE),]

#Get dates from 2001
stormTimeLine <- subset(stormTimeLine, as.Date("1/1/2001" , "%m/%d/%Y") < DATE)
#Delete irrelevant events
stormTimeLine <- subset(stormTimeLine, DMG != 0 )

type1_data <- stormTimeLine[stormTimeLine$EVTYPE == econ_top10_types[1],]
type2_data <- stormTimeLine[stormTimeLine$EVTYPE == econ_top10_types[2],]
type3_data <- stormTimeLine[stormTimeLine$EVTYPE == econ_top10_types[3],]

type1_data$CUMDMG <- cumsum(type1_data$DMG)
type2_data$CUMDMG <- cumsum(type2_data$DMG)
type3_data$CUMDMG <- cumsum(type3_data$DMG)

stormTimeLine <- rbind(type1_data[c('EVTYPE','CUMDMG','DATE')],
                       type2_data[c('EVTYPE','CUMDMG','DATE')],
                       type3_data[c('EVTYPE','CUMDMG','DATE')])


```

###Results

```{r Health}

health_top10

```

```{r HealthPlot}

ggplot(health_plot, aes(x=EVENT, y=CASUALITIES, fill=TYPE)) + 
    geom_bar(stat="identity") + 
    scale_fill_manual(values = c("red", "orange")) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    ggtitle("Health Damage") +
    ylab("Casualities") +    
    xlab("Event")

```

```{r property}

econ_top10

```


```{r HealthPlot}

ggplot(health_plot, aes(x=EVENT, y=CASUALITIES, fill=TYPE)) + 
    geom_bar(stat="identity") + 
    scale_fill_manual(values = c("red", "orange")) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    ggtitle("Health Damage") +
    ylab("Casualities") +    
    xlab("Event")

```


###Conclusions

Conclusions